---
title: "Unsupervised learning: PCA & CA"
params:
  answers: true
mainfont: Arial
fontsize: 12pt
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
    df_print: paged
    theme: paper
    # pandoc_args: --output=classification_2_answers.html
  pdf_document:
    toc: true
    toc_depth: 1
    latex_engine: xelatex
---

# Introduction
In this practical, we will learn how to use principal components analysis and correspondence analysis.

We will use the package `ca`. For this, you will probably need to `install.packages("ca")` before running the `library()` functions.

```{r packages, warning = FALSE, message = FALSE}
library(ISLR)
library(tidyverse)
library(ca)
```


```{r seed, include = FALSE}
set.seed(45)
```

# Principal components analysis

---

__Load the questionnaire dataset and explore it.__

---

```{r load, message = FALSE}
ques_df <- read_csv("data/questionnaire.csv")

summary(ques_df)

# a lot of likert scale variables!
```

 
---

__Split the dataset into two, one data frame with the `BIGF` questions (`bigf_df`), and one with the `empathy` questions (`empa_df`). Hint: use `contains()` in combination with `select()` for this!__

---

```{r split}

bigf_df <- ques_df %>% select(contains("bigf"))
empa_df <- ques_df %>% select(contains("empa"))

```

--- 

__Standardise the two datasets using the `scale()` function.__

---

```{r scale}
bigf_sc <- 
  bigf_df %>% 
  scale() %>% 
  as_tibble()

empa_sc <- 
  empa_df %>% 
  scale() %>% 
  as_tibble()


# optionally, we can also plot the scaled datasets to see what happened.
bubble_plot <- function(df) {
  df %>% 
    gather(key = Question, value = Answer) %>% 
    mutate(Question = str_to_lower(Question)) %>%
    ggplot(aes(x = Question, y = Answer)) + 
    geom_count(colour = "#00008B") + 
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
}

cowplot::plot_grid(
  bubble_plot(bigf_df) + ggtitle("bigf unscaled"),
  bubble_plot(bigf_sc) + ggtitle("bigf scaled"),
  bubble_plot(empa_df) + ggtitle("empa unscaled"),
  bubble_plot(empa_sc) + ggtitle("empa scaled")
)
```

---

__Use the `prcomp()` function to create a principal components analysis for each of the scaled datasets. Save the result as `pca_bigf` and pca_empa`.__

---

```{r pca}
# use prcomp here?
pca_bigf <- prcomp(bigf_sc)
pca_empa <- prcomp(empa_sc)
```

---

__Use the first 2 principal components of the bigf and empa datasets to 
---

```{r plot}

bind_cols(as_tibble(pca_bigf$x[, 1:2]), empa = pca_empa$x[, 1]) %>% 
  ggplot(aes(x = PC1, y = empa, colour = ques_df$Gender)) +
  geom_point() +
  theme_minimal() +
  scale_colour_viridis_d()

```

# Correspondence analysis

We've preprocessed a dataset from kaggle on song lyrics for the purpose of this practical. You can find the original dataset [here](https://www.kaggle.com/mousehead/songlyrics) or in the `data/` directory. If you want to know which preprocessing steps have been used and how it has been saved, you can take a look at the file `data/song_data_preproc.R`.

The `songs_ca` dataset is stored as a `.RData` file, a native file format from `R` which efficiently stores any `R` object. The `load()` function immediately loads the dataset `songs_ca` into your environment.

---

__Load the preprocessed `songs_ca` dataset into the environment from the `data/songs_ca.RData` file.__

---

```{r loadca}
load("data/songs_ca.RData")
```


---

__Use the `ca()` function from the `ca` package to create a correspondence analysis object.__

---

```{r ca1}
ca_mod <- ca(songs_ca)
```


---

__Use the `summary()` function on this object. What can you conclude about the first two inertias? What can you say about the word "love" in this dataset?__

---

```{r sumca}
summary(ca_mod)
```

---

__Recreate using `ggplot` the biplot that results from the `plot()` method on this object. Hint: for this, you can use the `rowcoord` and `colcoord` elements of the object.__

---

```{r ggca}
gg_ca <- 
  rbind(ca_mod$rowcoord[, 1:2], ca_mod$colcoord[, 1:2]) %>% 
  as_tibble() %>% 
  mutate(name = c(rownames(songs_ca), colnames(songs_ca)), 
         type = c(rep("row", 12), rep("col", 26)))

gg_ca %>%
  ggplot(aes(x = Dim1, y = Dim2, shape = type, colour = type, label = name)) +
  geom_point() +
  geom_text(hjust = 0, nudge_x = 0.05) +
  coord_fixed() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Dimension 1", y = "Dimension 2", 
       title = "CA of popular words in songs")

```

---

__What can you conclude about Exo and Janis Joplin? Can you come up with reasonable explanations for this?__

---

```{r ans}

# Exo is a k-pop band. Janis Joplin was born in 1943 (and a member of the 
# infamous 27 club), so her lyrics are from a different period than the other
# artists.

```

# Final assignment: High-dimensional PCA using SVD

This is an advanced assignment. You will have to figure out how to create PC scores from the output of a singular value decomposition.

Principal components analysis can also be used to generate a low-dimensional number of features from a high-dimensional (p > n) dataset. One area where high-dimensional data frequently occurs is in chemometrics, assessing the properties of materials using spectroscopy [(Wikipedia link)](https://en.wikipedia.org/w/index.php?title=Near-infrared_spectroscopy&oldid=865590066).

---

__Load the dataset "data/corn.RData" using the function `load()`.__

---

```{r loadcorn}
# Load the data. Source: http://www.eigenvector.com/data/Corn/index.html 
# converted from matlab to an R data object.
load("data/corn.RData")
```

The first four columns contain properties of the corn samples (80 corn samples were analysed) and the remaining 700 columns indicate the measured transmittance at different near-infrared wavelengths. You can find more information about this dataset at [the source website](http://www.eigenvector.com/data/Corn/index.html).

Here is a plot of wavelength versus transmittance, with one line for each of the 80 corn samples:

```{r cornplot}
t(corn[, -c(1:4)]) %>% 
  as_tibble %>% 
  gather(key = corn, value = signal) %>% 
  mutate(wavelength = rep(seq(1100, 2498, 2), 80)) %>% 
  ggplot(aes(x = wavelength, y = signal, colour = corn)) +
  geom_line() +
  theme_minimal() +
  scale_colour_viridis_d(guide = "none") +
  labs(x = "Wavelength (nm)",
       y = "Transmittance",
       title = "NIR Spectroscopy of 80 corn samples")
```


---

__Use the `svd()` function to run a principal components analysis on the spectroscopy part of the corn dataset. Save the PC scores and plot the first two principal components. Create four plots, each mapping one of the four properties to the `colour` aesthetic. Which of the properties relate most to the first two principal components? Base your answer on the plots only. Then, do the same thing for PCs 5 and 6.__

---


```{r svdpca}
# see https://stats.stackexchange.com/a/134283 for information.

# Perform svd-pca
x_scaled  <- scale(corn[, -c(1:4)], scale = FALSE)
svd_corn  <- svd(x_scaled)
pc_scores <- svd_corn$u %*% diag(svd_corn$d) 

# Plot first two principal components versus the corn properties 
ggcorn <-
  corn[, 1:4] %>%
  bind_cols(tibble(PC1 = pc_scores[, 1], 
                   PC2 = pc_scores[, 2], 
                   PC5 = pc_scores[, 5], 
                   PC6 = pc_scores[, 6]))

cowplot::plot_grid(
  ggcorn %>% 
    ggplot(aes(x = PC1, y = PC2, colour = Moisture)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c(),
  ggcorn %>% 
    ggplot(aes(x = PC1, y = PC2, colour = Oil)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c(),
  ggcorn %>% 
    ggplot(aes(x = PC1, y = PC2, colour = Protein)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c(),
  ggcorn %>% 
    ggplot(aes(x = PC1, y = PC2, colour = Starch)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c()
)

# Moisture and oil seem to have a strong relation with the first two PCs, 
# and their high values are on opposite sides.
# The high/low values for protein and starch seem more randomly 
# ordered in these plots. 

cowplot::plot_grid(
  ggcorn %>% 
    ggplot(aes(x = PC5, y = PC6, colour = Moisture)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c(),
  ggcorn %>% 
    ggplot(aes(x = PC5, y = PC6, colour = Oil)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c(),
  ggcorn %>% 
    ggplot(aes(x = PC5, y = PC6, colour = Protein)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c(),
  ggcorn %>% 
    ggplot(aes(x = PC5, y = PC6, colour = Starch)) +
    geom_point() +
    theme_minimal() + 
    scale_colour_viridis_c()
)

# Here, it's protein and starch that relate more to the PCs. They too are 
# opposites in these samples.
```
