---
title: "R Basics for DAV"
author: ""
date: ""
mainfont: Arial
fontsize: 12pt
urlcolor: blue
output: 
  pdf_document:
    toc: true
    toc_depth: 1
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 1
---

```{r load_packages, include = FALSE}
library(ISLR)
library(tidyverse)
library(haven)
```

# Loading data

For this practical, we are going to use a dataset from Kaggle - the Google play store apps data by user lava18. We have downloaded it into the data folder already from https://www.kaggle.com/lava18/google-play-store-apps.

Tidyverse contains many data loading functions in the packages `readr` (default file types) and `haven` (external file types such as from SPSS or Stata), each for their own file type. The most common file type is `csv`, which is what we use here.

> Use the function `read_csv()` to import the file "data/googleplaystore.csv" and store it in a variable called `apps`.

```{r readcsv}
apps <- read_csv("data/googleplaystore.csv")
```

These import functions are fast and safe: they display informative errors if anything goes wrong. In this case, we can also see what variable type each column gets.

> Use the function and `head()` to look at the first few rows of the `apps` dataset

```{r head}
head(apps)
```


